---
title: "HW1_Basic_Monte_Carlo_and_Bayesian_inference"
author: "Kiseok Lee"
date: "2022-01-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=11, fig.height=9,
                      error=TRUE, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE,
                      tidy.opts=list(width.cutoff=40),tidy=TRUE)
```

## HW1 Basic Monte Carlo method; Likelihood/Bayesian inference
Name: **Kiseok Lee** \
Date: 1/15/22 \
Class: **HGEN 486 Computational Biology**

```{r, echo=FALSE}
# libraries
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(vegan)
library(tidyverse)
library(magrittr)
library(readxl)
library(reshape2)
library(gtools)
library(devtools)
library(openxlsx)
library(ape)
library(stringr)
library(tidyr)
library(ggrepel)
library(ggpubr)

## theme for ggplot
mytheme <- theme_bw() + 
  theme(text = element_text(family="serif")) +
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=10, family="serif"))+
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank(),panel.background=element_blank(),panel.border=element_blank(),plot.background=element_blank())+
  theme(axis.ticks = element_line(size = 1.1))

mytheme_2d <- theme_bw() + 
  theme(text = element_text(family="serif")) +
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=13, family="serif"))+
  # theme(panel.grid.major = element_blank()) +
  # theme(panel.grid.minor = element_blank(),panel.background=element_blank(),plot.background=element_blank()) +
  theme(axis.ticks = element_line(size = 1.1))


# color collection
my_color_collection <- c(
  "#CBD588", "#5F7FC7", "orange", "#AD6F3B", "#673770", 
  "#D14285", "#652926", "#C84248", "#8569D5", "#5E738F",
  "#D1A33D", "#8A7C64", "#599861","#616163", "#FFCDB2",
  "#6D9F71", "#242F40",
  "#CCA43B", "#F92A82", "#ED7B84", "#7EB77F", 
  "#DEC4A1", "#E5D1D0", '#0E8482', '#C9DAEA', '#337357', 
  '#95C623', '#E55812', '#04471C', '#F2D7EE', '#D3BCC0', 
  '#A5668B', '#69306D', '#0E103D', '#1A535C', '#4ECDC4', 
  '#F7FFF7', '#FF6B6B', '#FFE66D', '#6699CC', '#FFF275', 
  '#FF8C42', '#FF3C38', '#A23E48', '#000000', '#CF5C36', 
  '#EEE5E9', '#7C7C7C', '#EFC88B', '#2E5266', '#6E8898', 
  '#9FB1BC', '#D3D0CB', '#E2C044', '#5BC0EB', '#FDE74C', 
  '#9BC53D', '#E55934', '#FA7921', "#CD9BCD", "#508578", "#DA5724")

# for git push, use this instead of using wflow_git_push()
# git push -u origin master (in the Git app / in the working directory)

# for making pdf file
library(rmarkdown)
# render("analysis/~~.Rmd", "pdf_document")

```

## A-2-c) Inverse Transform Sampling simulation
Veryfiy your previous answer by simulation. Specifically, using R to simulate 1000 values of U and compute 1000 corresponding values of - log (U). \
Plot the empirical cumulative distribution function (in R, on can use plot(ecdf(x))) for your resulting random deviates and show they are qualitatively similar to that expected for an exponential random variable by overlaying a plot of the expected cdf.


```{r}
# -log(U) sampling 1000 times
U = runif(1000)
x = -log(U)

# Plot the empirical cumulative distribution function
plot(ecdf(x), col='blue', lty = 1,pch=19,lwd=1, main="Empirical cumulative distribution function")

# Exponential random variable
set.seed(1)
k = rexp(1000, rate = 1) # rate is 1 for mean to be 1
lines(ecdf(k), col='red', type = "b",lty = 1, pch=19,lwd=1)

legend(2,0.2, legend=c('-log(U)', 'Exponential random variable'), col=c("Blue", "Red"), lty=1, cex=0.8, box.lty=0)

```

## Exercise: Likelihood Ratio: how big is convincing?
https://stephens999.github.io/fiveMinuteStats/LR_and_BF.html

## a. 
Write a function to simulate data for the medical screening example above. \
The function should take as input the proportion of individuals who have the disease, and the number of individuals to simulate. \
It should output a table, one row for each individual, with two columns: the first column (x) is the protein concentration for that individual, the second column (z) is an indicator of disease status (1=disease, 0=normal). \

First let's build the components of the function.
```{r}
# distribution protein concentration in blood of the diseased and healthy 
x = seq(0,10,length=100)
plot(x, dgamma(x,scale = 0.5,shape = 2), type="l", xlab="Protein concentration", ylab="density", main="protein concentration distribution in blood of the diseased and healthy")
lines(x, dgamma(x,scale = 1,shape = 2), type="l", col="red")
legend(4,0.6, legend=c('Healthy', 'Diseased'),
       col=c("Black", "Red"), lty=1, cex=0.8, box.lty=0)

# Input
# proportion of individuals who have the disease
p_dis = 0.1
# the number of individuals to simulate
n_ind = 1000

# Protein concentration random generation
set.seed(1) 

# healthy
healthy_protein_conc <- function(n){rgamma(n, scale = 0.5,shape = 2)}
healthy_protein_conc(n=1)

# diseased
diseased_protein_conc <- function(n){rgamma(n, scale = 1, shape = 2)}
diseased_protein_conc(n=1)

# Sample individual with diseased or no disease
z = sample(0:1, size=n_ind, prob=c(1 - p_dis, p_dis), replace=T)

# create x vector (protein concentration)
x = rep(-1, n_ind)

for (i in 1:n_ind){
  # print(i)
  if (z[i] == 0){  # healthy
    x[i] = healthy_protein_conc(n=1)
    # print(x[i])
  } else if (z[i] == 1){ # diseased
    x[i] = diseased_protein_conc(n=1)
    # print(x[i])
  } else{
    print("Error")
  }
}

head(x)

# bind columns
table_concentration <- cbind(x, z)
head(table_concentration)
```

Wrap it into a function
```{r}
## Wrapping into a function

#' @param p_dis proportion of individuals who have the disease
#' @param n_ind the number of individuals to simulate
simulate_disease <- function(p_dis, n_ind){
  # Sample individual with diseased or no disease
  z = sample(0:1, size=n_ind, prob=c(1 - p_dis, p_dis), replace=T)
  
  # create x vector (protein concentration)
  x = rep(-1, n_ind)
  
  for (i in 1:n_ind){
    # print(i)
    if (z[i] == 0){  # healthy
      x[i] = rgamma(1, scale = 0.5,shape = 2)
      # print(x[i])
    } else if (z[i] == 1){ # diseased
      x[i] = rgamma(1, scale = 1, shape = 2)
      # print(x[i])
    } else{
      print("Error")
    }
  }

  # bind columns
  table_concentration <- cbind(x, z)
  return(table_concentration)
}

# example

simulate_disease(p_dis=0.4, n_ind=10)


```


## b.
Write a function to compute the likelihood ratio in the medical screening example.

Construct the elements of the function
```{r}
# When z == 0 (healthy), likelihood ratio is
healthy_LR <- function(x){dgamma(x, scale = 0.5,shape = 2) / dgamma(x, scale = 1, shape = 2)}
# When z == 1, likelihood ratio is..
diseased_LR <- function(x){dgamma(x, scale = 1,shape = 2) / dgamma(x, scale = 0.5, shape = 2)}

## function to compute the likelihood ratio
# input will be the simulated data
vec_LR = rep(-1, dim(table_concentration)[1])  # empty likelihood ratio vector
for (i in 1:dim(table_concentration)[1]){
  # print(i)
  if (table_concentration[,"z"][i] == 0){  # healthy
    vec_LR[i] = healthy_LR(table_concentration[,"x"][i])
    # print(x[i])
  } else if (table_concentration[,"z"][i] == 1){ # diseased
    vec_LR[i] = diseased_LR(table_concentration[,"x"][i])
    # print(x[i])
  } else{
    print("Error")
  }
}

head(vec_LR)
# bind to the table as another column
table_LR <- cbind(table_concentration, LR = vec_LR)


```

Wrap into a function
```{r}
#' @table_concentration this is an simulated data input with column x and z. x is protein concentration and z is healthy or diseased index of 0 and 1.

calculate_LR <- function(table_concentration){
  # When z == 0 (healthy), likelihood ratio is
  healthy_LR <- function(x){dgamma(x, scale = 0.5,shape = 2) / dgamma(x, scale = 1, shape = 2)}
  # When z == 1, likelihood ratio is..
  diseased_LR <- function(x){dgamma(x, scale = 1,shape = 2) / dgamma(x, scale = 0.5, shape = 2)}
  
  ## function to compute the likelihood ratio
  # input will be the simulated data
  vec_LR = rep(-1, dim(table_concentration)[1])  # empty likelihood ratio vector
  for (i in 1:dim(table_concentration)[1]){
    # print(i)
    if (table_concentration[,"z"][i] == 0){  # healthy
      vec_LR[i] = healthy_LR(table_concentration[,"x"][i])
      # print(x[i])
    } else if (table_concentration[,"z"][i] == 1){ # diseased
      vec_LR[i] = diseased_LR(table_concentration[,"x"][i])
      # print(x[i])
    } else{
      print("Error")
    }
  }
  
  head(vec_LR)
  # bind to the table as another column
  table_LR <- cbind(table_concentration, LR = vec_LR)
  return(table_LR)
}

# example
calculate_LR(table_concentration) %>% head()

```

## b. 2nd trial
Because the simulations did not agree with the theoretical predictions. I'll reconstruct the function to trouble shoot.

```{r}
calculate_diseased_LR <- function(table_concentration){
  # only use the likelihood ratio where diseased likelihood is in the numerator.
  diseased_LR <- function(x){dgamma(x, scale = 1,shape = 2) / dgamma(x, scale = 0.5, shape = 2)}

  vec_LR = rep(-1, dim(table_concentration)[1])  # empty likelihood ratio vector
  for (i in 1:dim(table_concentration)[1]){
    vec_LR[i] = diseased_LR(table_concentration[,"x"][i])
  }
  
  head(vec_LR)
  # bind to the table as another column
  table_LR <- cbind(table_concentration, LR = vec_LR)
  return(table_LR)
}

# example
calculate_diseased_LR(table_concentration) %>% head()
```


## c.
Use the above functions to answer the following question by simulation. \
Suppose we screen a population of individuals, 20% of which are diseased, and compute the LR for each individual screened. 
Among individuals with an LR “near” c, what proportion are truly diseased? 
Denoting this proportion qD(c), make a plot of log10(c) [x axis] vs qD(c) [y axis], with c varying from 1/10 to 10 say (log10(c) varies from -1 to 1.) Or maybe a wider range if you like (the wider the range, the larger the simulation study you will need to get reliable results).

```{r}
# simulation
table_concentration1 <- simulate_disease(p_dis=0.2, n_ind=10000)
head(table_concentration1)

table_LR1 <- calculate_LR(table_concentration1)
head(table_LR1)
hist(table_LR1[,"LR"], breaks = 100)

# "near" definition
c = 5   # LR
eps = 0.5  # [LR - eps, LR + eps]
class(table_LR1)

# convert to dataframe
df_LR1 <- data.frame(table_LR1)

# Individuals with an LR “near” c
df_near <- df_LR1 %>% filter(LR <= c + eps & LR >= c - eps)
df_near %>% head()
# proportion are truly diseased?
qD <- sum(df_near$z)/dim(df_near)[1]

qD

```

Wrap in to qD function
```{r}
#' @param table_LR1 table of simulated x, z, LR
#' @param c LR
#' @param eps epsilon, range of + - c

calc_qD <- function(table_LR1, c, eps){
  # convert to dataframe
  df_LR1 <- data.frame(table_LR1)
  # Individuals with an LR “near” c
  df_near <- df_LR1 %>% filter(LR <= c + eps & LR >= c - eps)
  # proportion are truly diseased?
  qD <- sum(df_near$z)/dim(df_near)[1]
  
  return(qD)
}

```

make a plot of log10(c) [x axis] vs qD(c) [y axis], with c varying from 1/10 to 10 say (log10(c) varies from -1 to 1.)
```{r}
c <- seq(from = 1/10, to = 10, length.out = 1000)

## Let's say we define "near" relative to c. eps = c/100
# calculate qD for vector c
vec_qD <- c()
for (i in 1:length(c)){
  # print(i)
  vec_qD[i] = calc_qD(table_LR1, c[i], eps=c/100)
}

plot(log10(c), vec_qD, type="l", xlab="log10(c)", ylab="qD(c): proportion of truly diseased", main="Proportion are truly diseased (LR range: 0.1 < c < 10)")

## Let's say we define "near" in terms of absolute value. eps = 0.05
# calculate qD for vector c
vec_qD <- c()
for (i in 1:length(c)){
  # print(i)
  vec_qD[i] = calc_qD(table_LR1, c[i], eps=0.05)
}

plot(log10(c), vec_qD, type="l", xlab="log10(c)", ylab="qD(c): proportion of truly diseased", main="Proportion are truly diseased (LR range: 0.1 < c < 10)")


# Function
plot_qD <- function(table_LR1, c, eps){
  vec_qD <- c()
  for (i in 1:length(c)){
    # print(i)
    vec_qD[i] = calc_qD(table_LR1, c[i], eps)
  }
  
  plot(log10(c), vec_qD, type="l", xlab="log10(c)", ylab="qD(c): proportion of truly diseased", main=paste0("Proportion are truly diseased (eps= ",eps,")"))
}

plot_qD(table_LR1, c, eps=0.1)
plot_qD(table_LR1, c, eps=0.2)
plot_qD(table_LR1, c, eps=0.3)
plot_qD(table_LR1, c, eps=2)


```

## c. 2nd trial

Using the new LR. 
```{r}

table_concentration1 <- simulate_disease(p_dis=0.2, n_ind=10000)
table_LR1 <- calculate_diseased_LR(table_concentration1)

c <- seq(from = 1/10, to = 10, length.out = 1000)

plot_qD(table_LR1, c, eps=0.1)
plot_qD(table_LR1, c, eps=0.2)
plot_qD(table_LR1, c, eps=0.3)
plot_qD(table_LR1, c, eps=2)


```



## d.
Use the computations introduced in this vignette to compute the theoretical value for qD(c), and plot these on the same graph as your simulation results to demonstrate that your simulations match the theory. [It should provide a good agreement, provided your simulation is large enough.] \

First, let's get the theoretical value of qD(c), using the computation introduced in this vignette \
https://stephens999.github.io/fiveMinuteStats/LR_and_BF.html
```{r}
# computation introduced in this vignette
p_dis = 0.2 # proportion of diseased
c = 3 # c is LR
qD_vig = p_dis * c / ((1-p_dis) + p_dis * c)
qD_vig

calc_qD_vig <- function(c, p_dis){
  qD_vig = p_dis * c / ((1-p_dis) + p_dis * c)
  return(qD_vig)
}

# plot theoretical
c <- seq(from = 1/10, to = 10, length.out = 1000)
calc_qD_vig(c, p_dis = 0.2) %>% head()

plot(log10(c), calc_qD_vig(c, p_dis = 0.2), type="l", xlab="log10(c)", ylab="qD(c)", main="Theoretical value of qD(c)")

# overlay simulated qD above
vec_qD <- c()
for (i in 1:length(c)){
  # print(i)
  vec_qD[i] = calc_qD(table_LR1, c[i], eps=c[i]/10)
}

plot(log10(c), vec_qD, type="l", xlab="log10(c)", ylab="qD(c): proportion of truly diseased", main="Theoretical value of qD(c) vs. Simulated")

lines(log10(c), calc_qD_vig(c, p_dis = 0.2), col="blue")
legend(-0.9, 0.8, legend=c('Simulated', 'Theoretical'),
       col=c("Black", "Blue"), lty=1, cex=0.7, box.lty=0)


```
The theoretical prediction cannot capture the local increase and decrease at -0.5 < log10(c) < 0.

## e.
Repeat the above, but in the case where only 2% of the screened population are diseased.

```{r}
# simulation
table_concentration2 <- simulate_disease(p_dis=0.02, n_ind=10000)
head(table_concentration2)

table_LR2 <- calculate_LR(table_concentration2)
head(table_LR2)

# overlay simulated qD above
c <- seq(from = 1/10, to = 10, length.out = 1000)

vec_qD <- c()
for (i in 1:length(c)){
  # print(i)
  vec_qD[i] = calc_qD(table_LR2, c[i], eps=c[i]/10)
}

plot(log10(c), vec_qD, type="l", xlab="log10(c)", ylab="qD(c): proportion of truly diseased", main="Theoretical value of qD(c) vs. Simulated")

lines(log10(c), calc_qD_vig(c, p_dis = 0.02), col="blue")
legend(-0.9, 0.9, legend=c('Simulated', 'Theoretical'),
       col=c("Black", "Blue"), lty=1, cex=0.7, box.lty=0)


```

Try with larger c range
```{r}
c <- seq(from = 1/50, to = 50, length.out = 1000)

# overlay simulated qD above
vec_qD <- c()
for (i in 1:length(c)){
  # print(i)
  vec_qD[i] = calc_qD(table_LR2, c[i], eps=c[i]/10)
}

plot(log10(c), vec_qD, type="l", xlab="log10(c)", ylab="qD(c): proportion of truly diseased", main="Theoretical value of qD(c) vs. Simulated")

lines(log10(c), calc_qD_vig(c, p_dis = 0.02), col="blue")
legend(-0.9, 0.9, legend=c('Simulated', 'Theoretical'),
       col=c("Black", "Blue"), lty=1, cex=0.7, box.lty=0)

```

## e. 2nd trial

With the new LR function
```{r}
# simulation
table_concentration2 <- simulate_disease(p_dis=0.02, n_ind=10000)
head(table_concentration2)

table_LR2 <- calculate_diseased_LR(table_concentration2)
head(table_LR2)

# overlay simulated qD above
c <- seq(from = 1/10, to = 10, length.out = 1000)

vec_qD <- c()
for (i in 1:length(c)){
  # print(i)
  vec_qD[i] = calc_qD(table_LR2, c[i], eps=0.05)
}

plot(log10(c), vec_qD, type="l", xlab="log10(c)", ylab="qD(c): proportion of truly diseased", main="Theoretical value of qD(c) vs. Simulated")

lines(log10(c), calc_qD_vig(c, p_dis = 0.02), col="blue")
legend(-0.9, 0.9, legend=c('Simulated', 'Theoretical'),
       col=c("Black", "Blue"), lty=1, cex=0.7, box.lty=0)



```

Try with larger c range
```{r}
c <- seq(from = 1/50, to = 50, length.out = 1000)

# overlay simulated qD above
vec_qD <- c()
for (i in 1:length(c)){
  # print(i)
  vec_qD[i] = calc_qD(table_LR2, c[i], eps=c[i]/10)
}

plot(log10(c), vec_qD, type="l", xlab="log10(c)", ylab="qD(c): proportion of truly diseased", main="Theoretical value of qD(c) vs. Simulated")

lines(log10(c), calc_qD_vig(c, p_dis = 0.02), col="blue")
legend(-0.9, 0.9, legend=c('Simulated', 'Theoretical'),
       col=c("Black", "Blue"), lty=1, cex=0.7, box.lty=0)

```

Compared with disease proportion of 0.2, this 0.02 simulation result agrees less with the theoretical prediction.
