---
title: "HW3_Discrete_Markov_chains_Graphical_models"
author: "Kiseok Lee"
date: "2022-01-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=11, fig.height=9,
                      error=TRUE, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE,
                      tidy.opts=list(width.cutoff=40),tidy=TRUE)
```

## HW3 Discrete Markov chains Graphical models
Name: **Kiseok Lee** \
Date: 1/29/22 \
Class: **HGEN 486 Computational Biology**

```{r, echo=FALSE}
# libraries
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(vegan)
library(tidyverse)
library(magrittr)
library(readxl)
library(reshape2)
library(gtools)
library(devtools)
library(openxlsx)
library(ape)
library(stringr)
library(tidyr)
library(ggrepel)
library(ggpubr)

## theme for ggplot
mytheme <- theme_bw() + 
  theme(text = element_text(family="serif")) +
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=10, family="serif"))+
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank(),panel.background=element_blank(),panel.border=element_blank(),plot.background=element_blank())+
  theme(axis.ticks = element_line(size = 1.1))

mytheme_2d <- theme_bw() + 
  theme(text = element_text(family="serif")) +
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=13, family="serif"))+
  # theme(panel.grid.major = element_blank()) +
  # theme(panel.grid.minor = element_blank(),panel.background=element_blank(),plot.background=element_blank()) +
  theme(axis.ticks = element_line(size = 1.1))


# color collection
my_color_collection <- c(
  "#CBD588", "#5F7FC7", "orange", "#AD6F3B", "#673770", 
  "#D14285", "#652926", "#C84248", "#8569D5", "#5E738F",
  "#D1A33D", "#8A7C64", "#599861","#616163", "#FFCDB2",
  "#6D9F71", "#242F40",
  "#CCA43B", "#F92A82", "#ED7B84", "#7EB77F", 
  "#DEC4A1", "#E5D1D0", '#0E8482', '#C9DAEA', '#337357', 
  '#95C623', '#E55812', '#04471C', '#F2D7EE', '#D3BCC0', 
  '#A5668B', '#69306D', '#0E103D', '#1A535C', '#4ECDC4', 
  '#F7FFF7', '#FF6B6B', '#FFE66D', '#6699CC', '#FFF275', 
  '#FF8C42', '#FF3C38', '#A23E48', '#000000', '#CF5C36', 
  '#EEE5E9', '#7C7C7C', '#EFC88B', '#2E5266', '#6E8898', 
  '#9FB1BC', '#D3D0CB', '#E2C044', '#5BC0EB', '#FDE74C', 
  '#9BC53D', '#E55934', '#FA7921', "#CD9BCD", "#508578", "#DA5724")

# for git push, use this instead of using wflow_git_push()
# git push -u origin master (in the Git app / in the working directory)

# for making pdf file
library(rmarkdown)
# render("analysis/~~.Rmd", "pdf_document")

```

## A-a). Sequence of nucleotides
Assuming you look at a random location in the sequence, kilobases away from its starting point, calculate the probability of observing the sequence TATA under this model.  (Hint: Consider the sequence of elementary events that compose the event TATA and note that you are sampling from the stationary distribution). \

First, let's get the stationary distribution
```{r}

# probability transition matrix
P <- matrix(c(0.1,0.35,0.3,0.6,0.8,0.1,0.2,0.1,0.05,0.1,0.2,0.25,0.05,0.45,0.3,0.05),nrow=4)

# function from a vignette
# https://stephens999.github.io/fiveMinuteStats/stationary_distribution.html

states <- c("a","c","t","g")
stationary <- function(transition) {
  stopifnot(is.matrix(transition) &&
        nrow(transition)==ncol(transition) &&
            all(transition>=0 & transition<=1))
  p <- diag(nrow(transition)) - transition
  A <- rbind(t(p),
         rep(1, ncol(transition)))
  b <- c(rep(0, nrow(transition)),
         1)
  res <- qr.solve(A, b)
  names(res) <- paste0("state.", states[1:nrow(transition)])
  return(res)
}

stationary(P)

```

Then let's get the probability of TATA.
Pr(TATA) = pi(t) * Pt->a * Pa->t * Pt->a
Pr(TATA) = 0.00058295

```{r}
vec_pi <- stationary(P)
names(vec_pi) <- NULL
vec_pi[3] * P[3,1] * P[1,3] * P[3,1]
```

## A-b)

```{r}

# simulate discrete Markov chains
# source: https://stephens999.github.io/fiveMinuteStats/stationary_distribution.html
run.mc.sim <- function( P,   # probability transition matrix
                        num.iters=50, 
                        num.chains=150 )
  {
  
  # number of possible states
  num.states <- nrow(P)
  
  # states X_t for all chains
  states     <- matrix(NA, ncol=num.chains, nrow=num.iters)
  
  # probability vectors pi^n through time
  all_probs  <- matrix(NA, nrow=num.iters, ncol=num.states)
  
  # forces chains to start in state 1
  pi_0      <- c(1, rep(0, num.states-1))

  # initialize variables for first state 
  P_n           <- P
  all_probs[1,] <- pi_0
  states[1,]    <- 1

  for(t in 2:num.iters) {
    
    # pi^n for this iteration
    pi_n           <- pi_0 %*% P_n
    all_probs[t,]  <- pi_n
    
    for(chain_num in seq_len(num.chains)) {
      # probability vector to simulating next state 
      p                     <- P[ states[t-1,chain_num], ]
      states[t,chain_num]   <- which(rmultinom(1, 1, p) == 1)
    }
    
    # update probability transition matrix
    P_n           <- P_n %*% P
  }
  return(list(all.probs=all_probs, states=states))
}

# 10000 is burn-in, the next is the stationary phase
l_result <- run.mc.sim(P, num.iters=20000, num.chains=1000)
dim_states <- dim(l_result$states)
dim_states

# get the stationary phase (after the burn-in)
stationary_states <- l_result$states[10001:dim_states[1],]
dim(stationary_states)

# get every 100th base
seq(from=1, to=dim(stationary_states)[1],by=100)
states_100th <- stationary_states[seq(from=1, to=dim(stationary_states)[1],by=100),]
dim(states_100th)

# put all the state values into a vector (from a state matrix)
c(states_100th) %>% head()
# frequncy table
t_state <- table(c(states_100th)) / length(c(states_100th))
t_state
names(t_state) <- c("A","C","T","G")
df_state <- data.frame(t_state)

# plot
ggplot(df_state, aes(x=Var1, y=Freq)) +
  geom_bar(stat="identity", fill="maroon") +
  # scale_fill_manual(values=cols) +
  ylab("Probability of observing at stationary distribution \n") +
  xlab("\n Base (stationary phase)") +
  scale_y_continuous(limits=c(0,1),breaks=seq(0,1,0.1))+
  ggtitle("Probability in stationary distribution \n")+
  geom_text(aes(label = round(Freq,3)), size = 3, vjust = -1.5, family="serif", show.legend = FALSE, position = position_dodge(0.8))+
  mytheme_2d

vec_pi
```
The Monte Carlo estimate is very close to the exact solution.

## B-d) Simulation of graphical models
Show code for how in the software R, you would go about using the random number generators (rbinom, rpois) to simulate 1000 realizations for the following parameter values: p=0.5, n=500. Run the simulations and store the values for the following problems.

```{r}
p = 0.5
n = 500
r = 1000 # realizations

X1 <- rbinom(r, n, p)
N1 <- rpois(r, lambda = n)

X2 <- rbinom(r, size=N1, prob=X1/N1)
N2 <- rpois(r, lambda = N1)

X3 <- rbinom(r, size=N2, prob=X2/N2)
N3 <- rpois(r, lambda = N2)

X4 <- rbinom(r, size=N3, prob=X3/N3)
N4 <- rpois(r, lambda = N3)

```

## B-e) Monte carlo estimation
Monte Carlo estimate and the n*p (from derivation) are close.
```{r}
# Monte Carlo estimate of E[X2]
mean(X2)

# From derivation of E[X2]
n*p

# difference
mean(X2) - n*p
```

## B-f) MC estimation of P(X4>X1)
It is approximately 0.5.
E[X1]=E[X2]=E[X3]=E[X4]. Therefore, P(X4 > X1) will be 0.5.
```{r}
# Monte Carlo estimation of P(X4>X1)
mean(X4>X1)

```




