---
title: "HW5_HMM_Poisson_processes"
author: "Kiseok Lee"
date: "2022-02-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=11, fig.height=9,
                      error=TRUE, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE,
                      tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## HW5 HMMs (Hidden Markov models), Poisson processes
Name: **Kiseok Lee** \
Date: 2/24/22 \
Class: **HGEN 486 Computational Biology**


```{r, echo=FALSE}
# libraries
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(vegan)
library(tidyverse)
library(magrittr)
library(readxl)
library(reshape2)
library(gtools)
library(devtools)
library(openxlsx)
library(ape)
library(stringr)
library(tidyr)
library(ggrepel)
library(ggpubr)

## theme for ggplot
mytheme <- theme_bw() + 
  theme(text = element_text(family="serif")) +
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=10, family="serif"))+
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank(),panel.background=element_blank(),panel.border=element_blank(),plot.background=element_blank())+
  theme(axis.ticks = element_line(size = 1.1))

mytheme_2d <- theme_bw() + 
  theme(text = element_text(family="serif")) +
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=13, family="serif"))+
  # theme(panel.grid.major = element_blank()) +
  # theme(panel.grid.minor = element_blank(),panel.background=element_blank(),plot.background=element_blank()) +
  theme(axis.ticks = element_line(size = 1.1))


# color collection
my_color_collection <- c(
  "#CBD588", "#5F7FC7", "orange", "#AD6F3B", "#673770", 
  "#D14285", "#652926", "#C84248", "#8569D5", "#5E738F",
  "#D1A33D", "#8A7C64", "#599861","#616163", "#FFCDB2",
  "#6D9F71", "#242F40",
  "#CCA43B", "#F92A82", "#ED7B84", "#7EB77F", 
  "#DEC4A1", "#E5D1D0", '#0E8482', '#C9DAEA', '#337357', 
  '#95C623', '#E55812', '#04471C', '#F2D7EE', '#D3BCC0', 
  '#A5668B', '#69306D', '#0E103D', '#1A535C', '#4ECDC4', 
  '#F7FFF7', '#FF6B6B', '#FFE66D', '#6699CC', '#FFF275', 
  '#FF8C42', '#FF3C38', '#A23E48', '#000000', '#CF5C36', 
  '#EEE5E9', '#7C7C7C', '#EFC88B', '#2E5266', '#6E8898', 
  '#9FB1BC', '#D3D0CB', '#E2C044', '#5BC0EB', '#FDE74C', 
  '#9BC53D', '#E55934', '#FA7921', "#CD9BCD", "#508578", "#DA5724")

# for git push, use this instead of using wflow_git_push()
# git push -u origin master (in the Git app / in the working directory)

# for making pdf file
library(rmarkdown)
# render("analysis/~~.Rmd", "pdf_document")

```

## A. HMM: Pott's data on chromatin accessibility
- Pott's data introduction: http://stephens999.github.io/hgen48600/potts_1.html \
- Data here: https://github.com/stephens999/hgen48600/tree/master/data/potts \

## A-a) Implement forward algorithm
Implement the forward algorithm for an HMM and using these parameters, compute the probability of the observed Potts data. Spot-check that these are good MLE values by perturbing the values by a few percent of their values and seeing that the likelihood does not increase (or if it does, that the increase is not substantial!)
```{r}
# Input data 
setwd("C:/Users/there/Desktop/Computational_Biology_2022_Winter/fiveMinuteStats")
df_pott <- read.csv("data/Pott_methylation.tsv", sep = "\t", header=F, stringsAsFactors = FALSE)
head(df_pott)

# binarized data
m = 1*(df_pott$V7 > 0)
X = m + 1 # in order to put into matrix row index

# Transition Probability
mat_P = rbind(c(0.957, 1 - 0.957), c(1 - 0.58, 0.58))

# Emission probability
mat_E = rbind(c(1 - 0.067, 0.067), c(1 - 0.451, 0.451))

# prior distribution
pi = c(0.5, 0.5)

# Alpha
T = dim(df_pott)[1]
K = 2
mat_alpha = matrix(nrow = T,ncol=K)

# Initialize alpha[1,]
mat_alpha[1,] = pi * mat_E[,X[1]]
# mat_alpha[1,] = mat_alpha[1,] / sum(mat_alpha[1,])
head(mat_alpha)

# Forward algorithm
for(t in 1:(T-2)){
  mat_alpha[t+1,] = (mat_alpha[t,] %*% mat_P) * mat_E[,X[t+1]]
  # renormalize?
  mat_alpha[t+1,] = mat_alpha[t+1,] / sum(mat_alpha[t+1,])

}

head(mat_alpha)
tail(mat_alpha)

# calculating the last step
mat_alpha[T,] = (mat_alpha[T-1,] %*% mat_P) * mat_E[,X[T]]

# probability
sum(mat_alpha[T,])

```

Applying small perturbation
```{r}
# make it into a function


perturb_forward_HMM <- function(perturb=0.1){
  # Transition Probability
  mat_P = rbind(c(0.957*(1-perturb), 1 - 0.957*(1-perturb)), c(1 - 0.58*(1-perturb), 0.58*(1-perturb)))
  
  # Emission probability
  mat_E = rbind(c(1 - 0.067*(1-perturb), 0.067*(1-perturb)), c(1 - 0.451*(1-perturb), 0.451*(1-perturb)))
  # Alpha
  T = dim(df_pott)[1]
  K = 2
  mat_alpha = matrix(nrow = T,ncol=K)
  
  # Initialize alpha[1,]
  mat_alpha[1,] = pi * mat_E[,X[1]]
  # mat_alpha[1,] = mat_alpha[1,] / sum(mat_alpha[1,])
  head(mat_alpha)
  
  # Forward algorithm
  for(t in 1:(T-2)){
    mat_alpha[t+1,] = (mat_alpha[t,] %*% mat_P) * mat_E[,X[t+1]]
    # renormalize?
    mat_alpha[t+1,] = mat_alpha[t+1,] / sum(mat_alpha[t+1,])
  
  }
  
  # calculating the last step
  mat_alpha[T,] = (mat_alpha[T-1,] %*% mat_P) * mat_E[,X[T]]

  return(sum(mat_alpha[T,]))
}

perturb_forward_HMM(perturb=0)
perturb_forward_HMM(perturb=0.01)
perturb_forward_HMM(perturb=0.1)
perturb_forward_HMM(perturb=0.3)
perturb_forward_HMM(perturb=0.5)
perturb_forward_HMM(perturb=0.9)
```

The perturbation does not change the probability. However, I've discovered that perturbation by 90% actually increases the likelihood. This may be due to the symmetric scheme of perturbation

## A-b)
Using the parameter values specified above, provide a maximum a posteriori decoding of the sequence in terms of where it is "open" and "closed". Do so by implementing the Viterbi algorithm. 

```{r}




```





