---
title: "HW7_Brownian_Motion_Gaussian_process"
author: "Kiseok Lee"
date: "2022-03-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


```{r global_options, include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=11, fig.height=9,
                      error=TRUE, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE,
                      tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## HW7 Brownian Motion, Gaussian processes
Name: **Kiseok Lee** \
Date: 3/5/22 \
Class: **HGEN 486 Computational Biology**


```{r, echo=FALSE}
# libraries
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(vegan)
library(tidyverse)
library(magrittr)
library(readxl)
library(reshape2)
library(gtools)
library(devtools)
library(openxlsx)
library(ape)
library(stringr)
library(tidyr)
library(ggrepel)
library(ggpubr)

## theme for ggplot
mytheme <- theme_bw() + 
  theme(text = element_text(family="serif")) +
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=10, family="serif"))+
  theme(panel.grid.major = element_blank()) +
  theme(panel.grid.minor = element_blank(),panel.background=element_blank(),panel.border=element_blank(),plot.background=element_blank())+
  theme(axis.ticks = element_line(size = 1.1))

mytheme_2d <- theme_bw() + 
  theme(text = element_text(family="serif")) +
  theme(plot.title = element_text(size = 19,hjust = 0.5, family="serif")) + 
  theme(axis.title.x = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.title.y = element_text(size = 17,hjust = 0.5, family="serif")) + 
  theme(axis.text.x = element_text(hjust = 0.5, vjust=0.3,size=13, family="serif"))+
  theme(axis.text.y = element_text(size=13, family="serif"))+
  # theme(panel.grid.major = element_blank()) +
  # theme(panel.grid.minor = element_blank(),panel.background=element_blank(),plot.background=element_blank()) +
  theme(axis.ticks = element_line(size = 1.1))


# color collection
my_color_collection <- c(
  "#CBD588", "#5F7FC7", "orange", "#AD6F3B", "#673770", 
  "#D14285", "#652926", "#C84248", "#8569D5", "#5E738F",
  "#D1A33D", "#8A7C64", "#599861","#616163", "#FFCDB2",
  "#6D9F71", "#242F40",
  "#CCA43B", "#F92A82", "#ED7B84", "#7EB77F", 
  "#DEC4A1", "#E5D1D0", '#0E8482', '#C9DAEA', '#337357', 
  '#95C623', '#E55812', '#04471C', '#F2D7EE', '#D3BCC0', 
  '#A5668B', '#69306D', '#0E103D', '#1A535C', '#4ECDC4', 
  '#F7FFF7', '#FF6B6B', '#FFE66D', '#6699CC', '#FFF275', 
  '#FF8C42', '#FF3C38', '#A23E48', '#000000', '#CF5C36', 
  '#EEE5E9', '#7C7C7C', '#EFC88B', '#2E5266', '#6E8898', 
  '#9FB1BC', '#D3D0CB', '#E2C044', '#5BC0EB', '#FDE74C', 
  '#9BC53D', '#E55934', '#FA7921', "#CD9BCD", "#508578", "#DA5724")

# for git push, use this instead of using wflow_git_push()
# git push -u origin master (in the Git app / in the working directory)

# for making pdf file
library(rmarkdown)
# render("analysis/~~.Rmd", "pdf_document")

```

## A. Brownian Motion

## A-a) Standard Brownian motion

$$ X_{t+\mu} | X_t \sim N(0, \mu) $$

Let's simulate this standard Brownian motion and plot
```{r}

simulate_standard_Brownian = function(length=1000){
  t = c(0, runif(999, min=0, max=1))
  # increasing order
  t = t[order(t)]
  
  # Now Brownian motion
  X = rep(0,length)
  X[1] = 0  # X(0) = 0
  for(k in 2:length){
    # print(k)
    X[k]= X[k-1] + rnorm(1, mean = 0, sd = t[k] - t[k-1])  
  }
  return(list(X=X, t=t))
}

set.seed(100)
standard_sim <- simulate_standard_Brownian(length=1000)

plot(standard_sim$t, standard_sim$X, xlab = "t", ylab="X(t)", main = paste0("Standard Brownian motion"), col = "royalblue")

```


## A-b) Scaled Brownian motion
variance parameter $$\sigma ^ 2 $$

Markov chain
$$ X_{t+\mu} | X_t \sim N(0, \mu \sigma ^ 2) $$

likelihood of data given $\sigma where standard deviation of the normal distribution is
$$ \sigma \sqrt{t_{k} - t_{k-1}} $$

$$ P(X | \sigma) = \prod_{i=2}^k  \frac{1}{\sigma\sqrt{2\pi(t_i-t_{i-1})}} e^{-\frac{1}{2}({\frac{X_i - X_{i-1}}{\sigma \sqrt{(t_i-t_{i-1})}}})^2} $$
log-likelihood

$$ log P(X | \sigma)= \sum_{i=2}^k ( -\frac{1}{2}({\frac{X_i - X_{i-1}}{\sigma \sqrt{(t_i-t_{i-1})}}})^2 - log(\sigma\sqrt{2\pi(t_i-t_{i-1})}) )  $$

Plot the log-likelihood as a function of $\sigma

```{r}


calculate_loglikelihood <- function(sigma = 1, standard_sim){
  vec_ll <- c() # vector to store log-likelihood
  X = standard_sim$X
  t = standard_sim$t
  
  # -0.5 * ( (X[2] - X[1]) / (sigma*(t[2]-t[1])))^2 - log( sigma * sqrt(2*pi* (t[2]-t[1]) ))
  # loop
  for (i in 2:length(t)){
    ll <- -0.5 * ( (X[i] - X[i-1]) / (sigma*sqrt(t[i]-t[i-1]) ) )^2 - log( sigma * sqrt(2*pi* (t[i]-t[i-1]) ))
    # print(paste0("front: ", -0.5 * ( (X[i] - X[i-1]) / (sigma*sqrt(t[i]-t[i-1]) ) )^2))
    # print(paste0("back: ", - log( sigma * sqrt(2*pi* (t[i]-t[i-1]) ))))
    # print(paste0("back inside log: ", ( sigma * sqrt(2*pi* (t[i]-t[i-1]) ))))
    # print(ll)
    vec_ll <- c(vec_ll, ll)
  }
  # print(length(vec_ll))
  # print(sum(vec_ll))
  return(sum(vec_ll))
}

# when sigma is 1
calculate_loglikelihood(sigma = 1, standard_sim)

# plot for sigma values between 0 to 1
vec_loglikelihood <- c()
vec_sigma <- seq(0, 1, length.out = 1000)[-1]
for (sig in vec_sigma){ 
  # calculate_loglikelihood(sigma = sig, standard_sim)
  vec_loglikelihood <- c(vec_loglikelihood, calculate_loglikelihood(sigma = sig, standard_sim))
}

# plot
plot(vec_sigma, vec_loglikelihood, xlab = "Sigma", ylab="Log-likelihood(sigma)", main = paste0("log-likelihood function"), col = "royalblue")

```

Let's magnify the area between 0 and 0.1

```{r}
# plot for sigma values between 0 to 0.1
vec_loglikelihood <- c()
vec_sigma <- seq(0, 0.1, length.out = 1000)[-1]
for (sig in vec_sigma){ 
  # calculate_loglikelihood(sigma = sig, standard_sim)
  vec_loglikelihood <- c(vec_loglikelihood, calculate_loglikelihood(sigma = sig, standard_sim))
}

# plot
plot(vec_sigma, vec_loglikelihood, xlab = "Sigma", ylab="Log-likelihood(sigma)", main = paste0("log-likelihood function"), col = "royalblue")

# find maximum
max(vec_loglikelihood)
vec_sigma[which.max(vec_loglikelihood)] # 0.03303303
```

Let's magnify the area between 0.001 and 1

```{r}
# plot for sigma values between 0.001 to 1
vec_loglikelihood <- c()
vec_sigma <- seq(0.001, 1, length.out = 1000)[-1]
for (sig in vec_sigma){ 
  # calculate_loglikelihood(sigma = sig, standard_sim)
  vec_loglikelihood <- c(vec_loglikelihood, calculate_loglikelihood(sigma = sig, standard_sim))
}

# plot
plot(vec_sigma, vec_loglikelihood, xlab = "Sigma", ylab="Log-likelihood(sigma)", main = paste0("log-likelihood function"), col = "royalblue")

# find maximum
max(vec_loglikelihood)
vec_sigma[which.max(vec_loglikelihood)] # 0.033
```

The maximum seems to lie between 0.01 and 0.05. Let's zoom in between 0.015 and 0.05


```{r}
# plot for sigma values between 0.015 to 0.05
vec_loglikelihood <- c()
vec_sigma <- seq(0.015, 0.05, length.out = 1000)[-1]
for (sig in vec_sigma){ 
  # calculate_loglikelihood(sigma = sig, standard_sim)
  vec_loglikelihood <- c(vec_loglikelihood, calculate_loglikelihood(sigma = sig, standard_sim))
}

# plot
plot(vec_sigma, vec_loglikelihood, xlab = "Sigma", ylab="Log-likelihood(sigma)", main = paste0("log-likelihood function"), col = "royalblue")

# find maximum
max(vec_loglikelihood)
vec_sigma[which.max(vec_loglikelihood)] # 0.03297297
```

Maximum log-likelihood is acheived at sigma of 0.033

## A-c) Brownian motion in the tree
Consider the tree in Figure 1 of the Mau et al paper http://www.ncbi.nlm.nih.gov/pubmed/11318142 with leafs labelled 1 to 7.

Brownian motion
$$ X_{5}^i | X_{6}^i \sim N(0, t_6) $$

$$ X_{4}^i | X_{5}^i \sim N(0, t_5) $$
$$ X_{3}^i | X_{5}^i \sim N(0, t_4+t_5) $$

$$ X_{2}^i | X_{3}^i \sim N(0, t_3) $$
$$ X_{1}^i | X_{4}^i \sim N(0, t_4+t_3+t_2) $$

$$ X_{1}^L | X_{4}^i \sim N(0, t_4+t_3+t_2+t_1) $$

$$ X_{4}^L | X_{1}^i \sim N(0, t_1) $$

$$ X_{7}^L | X_{1}^i \sim N(0, t_1) $$
$$ X_{2}^L | X_{3}^i \sim N(0, t_3+t_2+t_1) $$

$$ X_{3}^L | X_{2}^i \sim N(0, t_1) $$
$$ X_{6}^L | X_{2}^i \sim N(0, t_1) $$

$$ X_{5}^L | X_{6}^i \sim N(0, t_6+t_5+t_4+t_3+t_2+t_1) $$
## A-c-1) Simulated value of $X^i$, $X^L$ given the coalescent times (t1, t2, ..., t6)

```{r}

set.seed(1)
t = rnorm(6, mean = 1, sd = 0.5) 

xi6 = 0
# X_{5}^i | X_{6}^i \sim N(0, t_6) 
xi5 = xi6 + rnorm(1, mean = 0, sd = sqrt(t[6]))
# X_{4}^i | X_{5}^i \sim N(0, t_5)
xi4 = xi5 + rnorm(1, mean = 0, sd = sqrt(t[5]))
# X_{3}^i | X_{5}^i \sim N(0, t_4+t_5)
xi3 = xi5 + rnorm(1, mean = 0, sd = sqrt(t[5]+t[4]))
# X_{2}^i | X_{3}^i \sim N(0, t_3)
xi2 = xi3 + rnorm(1, mean = 0, sd = sqrt(t[3]))
# X_{1}^i | X_{4}^i \sim N(0, t_4+t_3+t_2) 
xi1 = xi4 + rnorm(1, mean = 0, sd = sqrt(t[4]+t[3]+t[2]))
# X_{1}^L | X_{4}^i \sim N(0, t_4+t_3+t_2+t_1) 
xL1 = xi4 + rnorm(1, mean = 0, sd = sqrt(t[4]+t[3]+t[2]+t[1]))
# X_{4}^L | X_{1}^i \sim N(0, t_1) 
xL4 = xi1 + rnorm(1, mean = 0, sd = sqrt(t[1]))
# X_{7}^L | X_{1}^i \sim N(0, t_1) 
xL7 = xi1 + rnorm(1, mean = 0, sd = sqrt(t[1]))
# X_{2}^L | X_{3}^i \sim N(0, t_3+t_2+t_1) 
xL2 = xi3 + rnorm(1, mean = 0, sd = sqrt(t[3]+t[2]+t[1]))
# X_{3}^L | X_{2}^i \sim N(0, t_1) 
xL3 = xi2 + rnorm(1, mean = 0, sd = sqrt(t[1]))
# X_{6}^L | X_{2}^i \sim N(0, t_1) 
xL6 = xi2 + rnorm(1, mean = 0, sd = sqrt(t[1]))
# X_{5}^L | X_{6}^i \sim N(0, t_6+t_5+t_4+t_3+t_2+t_1) 
xL5 = xi6 + rnorm(1, mean = 0, sd = sqrt(t[6]+t[5]+t[4]+t[3]+t[2]+t[1]))

vec_x = c(xi6, xi5, xi4, xi3, xi2, xi1, xL1, xL4, xL7, xL2, xL3, xL6, xL5)

vec_x

# write it into function
simulate_xi_xL <- function(t){
  xi6 = 0
  # X_{5}^i | X_{6}^i \sim N(0, t_6) 
  xi5 = xi6 + rnorm(1, mean = 0, sd = sqrt(t[6]))
  # X_{4}^i | X_{5}^i \sim N(0, t_5)
  xi4 = xi5 + rnorm(1, mean = 0, sd = sqrt(t[5]))
  # X_{3}^i | X_{5}^i \sim N(0, t_4+t_5)
  xi3 = xi5 + rnorm(1, mean = 0, sd = sqrt(t[5]+t[4]))
  # X_{2}^i | X_{3}^i \sim N(0, t_3)
  xi2 = xi3 + rnorm(1, mean = 0, sd = sqrt(t[3]))
  # X_{1}^i | X_{4}^i \sim N(0, t_4+t_3+t_2) 
  xi1 = xi4 + rnorm(1, mean = 0, sd = sqrt(t[4]+t[3]+t[2]))
  # X_{1}^L | X_{4}^i \sim N(0, t_4+t_3+t_2+t_1) 
  xL1 = xi4 + rnorm(1, mean = 0, sd = sqrt(t[4]+t[3]+t[2]+t[1]))
  # X_{4}^L | X_{1}^i \sim N(0, t_1) 
  xL4 = xi1 + rnorm(1, mean = 0, sd = sqrt(t[1]))
  # X_{7}^L | X_{1}^i \sim N(0, t_1) 
  xL7 = xi1 + rnorm(1, mean = 0, sd = sqrt(t[1]))
  # X_{2}^L | X_{3}^i \sim N(0, t_3+t_2+t_1) 
  xL2 = xi3 + rnorm(1, mean = 0, sd = sqrt(t[3]+t[2]+t[1]))
  # X_{3}^L | X_{2}^i \sim N(0, t_1) 
  xL3 = xi2 + rnorm(1, mean = 0, sd = sqrt(t[1]))
  # X_{6}^L | X_{2}^i \sim N(0, t_1) 
  xL6 = xi2 + rnorm(1, mean = 0, sd = sqrt(t[1]))
  # X_{5}^L | X_{6}^i \sim N(0, t_6+t_5+t_4+t_3+t_2+t_1) 
  xL5 = xi6 + rnorm(1, mean = 0, sd = sqrt(t[6]+t[5]+t[4]+t[3]+t[2]+t[1]))
  
  vec_x = c(xi6, xi5, xi4, xi3, xi2, xi1, xL1, xL4, xL7, xL2, xL3, xL6, xL5)
  
  return(vec_x)
}

```

## A-c-2) assume tj=1 and compute covariance matrix for x

Here, we need to create matrix A.
Let's assign z1, z2, z3, z4, z5, ..., z11, z12 as links that connect nodes.
z0: root
z1: connecting xi6, xi5 \
z2: connecting xi5, xi4 \
z3: connecting xi5, xi3 \
z4: connecting xi3, xi2 \
z5: connecting xi4, xi1 \
z6: connecting xi4, xL1 \
z7: connecting xi1, xL4 \
z8: connecting xi1, xL7 \
z9: connecting xi3, xL2 \
z10: connecting xi2, xL3 \
z11: connecting xi2, xL6 \
z12: connecting xi6, xL5 \

Then, each row of A can be constructed based on the linear combination of links that each node can be reached from root (xi6)
```{r}
t = rep(1, 6)
simulate_xi_xL(t)

# length of link z (squre root of the time)
length_z = sqrt(c(1, 1, 1, 2, 1, 3, 4, 1, 1, 3, 2, 2, 6))

# xi6
row1 = c(1, 0,0,0,0,0, 0,0,0,0,0, 0,0)
row1 = row1 * length_z

# xi5
row2 = c(1, 1,0,0,0,0, 0,0,0,0,0, 0,0)
row2 = row2 * length_z

# xi4
row3 = c(1, 1,1,0,0,0, 0,0,0,0,0, 0,0)
row3 = row3 * length_z

# xi3
row4 = c(1, 1,0,1,0,0, 0,0,0,0,0, 0,0)
row4 = row4 * length_z

# xi2
row5 = c(1, 1,0,1,1,0, 0,0,0,0,0, 0,0)
row5 = row5 * length_z

# xi1
row6 = c(1, 1,1,0,0,1, 0,0,0,0,0, 0,0)
row6 = row6 * length_z

# xL1
row7 = c(1, 1,1,0,0,0, 1,0,0,0,0, 0,0)
row7 = row7 * length_z

# xL4
row8 = c(1, 1,1,0,0,1, 0,1,0,0,0, 0,0)
row8 = row8 * length_z

# xL7
row9 = c(1, 1,1,0,0,1, 0,0,1,0,0, 0,0)
row9 = row9 * length_z

# xL2
row10 = c(1, 1,0,1,0,0, 0,0,0,1,0, 0,0)
row10 = row10 * length_z

# xL3
row11 = c(1, 1,0,1,1,0, 0,0,0,0,1, 0,0)
row11 = row11 * length_z

# xL6
row12 = c(1, 1,0,1,1,0, 0,0,0,0,0, 1,0)
row12 = row12 * length_z

# xL5
row13 = c(1, 0,0,0,0,0, 0,0,0,0,0, 0,1)
row13 = row13 * length_z

mat_A = rbind(row1, row2, row3, row4, row5, row6, row7, row8, row9, row10, row11, row12, row13)

mat_covariance = mat_A %*% t(mat_A)
mat_covariance
```

## A-c-3) Confirm that my simulation agrees with covariance matrix

```{r}
# simulate 10000 times
t = rep(1, 6)

mat_data <- simulate_xi_xL(t)

for (i in 1:9999){
  mat_data <- rbind(mat_data, simulate_xi_xL(t))
}

dim(mat_data)
colnames(mat_data) = c("xi6", "xi5", "xi4", "xi3", "xi2", "xi1", "xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5")

cov(mat_data)

# covariance matrix  with simulation
library(lattice)
levelplot(cov(mat_data), main = "simulation (Problem A-c-1)")
a <- levelplot(cov(mat_data), main = "simulation (Problem A-c-1)")

# covariance matrix with logical structuring of z
# get matrix from problem A-c-2)
colnames(mat_covariance) = c("xi6", "xi5", "xi4", "xi3", "xi2", "xi1", "xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5")
rownames(mat_covariance) = c("xi6", "xi5", "xi4", "xi3", "xi2", "xi1", "xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5")
levelplot(mat_covariance, main = "logical structuring of z (Problem A-c-2))")
b <- levelplot(mat_covariance, main = "logical structuring of z (Problem A-c-2))")

# plot and compare
library(gridExtra)
grid.arrange(a, b, ncol=2)
```

They roughly agree.

## A-c-4) Precision matrix

Calculate the precision matrix with the covariance matrix from problem A-c-2
```{r}
library(matlib)
# covariance matrix from the problem A-c-2
mat_Omega <- inv(mat_covariance)
colnames(mat_Omega) = c("xi6", "xi5", "xi4", "xi3", "xi2", "xi1", "xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5")
rownames(mat_Omega) = c("xi6", "xi5", "xi4", "xi3", "xi2", "xi1", "xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5")

levelplot(mat_Omega, main = "Heatmap of Precision matrix (total)")
p1 <- levelplot(mat_Omega, main = "Heatmap of Precision matrix (total)")

# find covariance matrix for xi
cov_xi <- mat_covariance[c("xi6","xi5", "xi4", "xi3", "xi2", "xi1"), c("xi6","xi5", "xi4", "xi3", "xi2", "xi1")]
cov_xi
# inverse
inv(cov_xi)
levelplot(cov_xi, main = "Heatmap of Precision matrix of Xi")
p2 <- levelplot(cov_xi, main = "Heatmap of Precision matrix of Xi")


# find covariance matrix for xL
cov_xL <- mat_covariance[c("xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5"), c("xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5")]
cov_xL
# inverse
inv(cov_xL)
levelplot(cov_xL, main = "Heatmap of Precision matrix of X L")
p3 <- levelplot(cov_xL, main = "Heatmap of Precision matrix of X L")

# plot simultaneously
library(gridExtra)
grid.arrange(p1, p2, p3, ncol=3)

```

Calculate the precision matrix with the covariance matrix from problem A-c-3 (simulation-based)
```{r}
library(matlib)
# covariance matrix from the problem A-c-3
library(pracma)
mat_covariance_sim <- cov(mat_data)
mat_Omega <- pinv(mat_covariance_sim) # use pseudo inverse, due to the matrix being singular
colnames(mat_Omega) = c("xi6", "xi5", "xi4", "xi3", "xi2", "xi1", "xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5")
rownames(mat_Omega) = c("xi6", "xi5", "xi4", "xi3", "xi2", "xi1", "xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5")

levelplot(mat_Omega, main = "Heatmap of Precision matrix (total, simulation)")
p1 <- levelplot(mat_Omega, main = "Heatmap of Precision matrix (total, simulation)")

# find covariance matrix for xi
cov_xi <- mat_covariance_sim[c("xi6","xi5", "xi4", "xi3", "xi2", "xi1"), c("xi6","xi5", "xi4", "xi3", "xi2", "xi1")]
cov_xi
# inverse
pinv(cov_xi)
levelplot(cov_xi, main = "Heatmap of Precision matrix of Xi(simulation)")
p2 <- levelplot(cov_xi, main = "Heatmap of Precision matrix of Xi(simulation)")


# find covariance matrix for xL
cov_xL <- mat_covariance_sim[c("xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5"), c("xL1", "xL4", "xL7", "xL2", "xL3", "xL6", "xL5")]
cov_xL
# inverse
inv(cov_xL)
levelplot(cov_xL, main = "Heatmap of Precision matrix of X L(simulation)")
p3 <- levelplot(cov_xL, main = "Heatmap of Precision matrix of X L(simulation)")

# plot simultaneously
library(gridExtra)
grid.arrange(p1, p2, p3, ncol=3)

```

Both the simulation-based and the graphical model-based covariance matrix produces similar precision matrix. \

Component of precision matrix is zero if and only if $X_i$ and $X_j$ are conditionally independent given all other coordinates of X.\
Therefore, sparsity patterns depends on the conditional dependence/independence of Xs. \

(1) When we look at the "Precision matrix of $X^i$", we can see that the sparsity pattern agrees with the Graphical model. In the heatmap, Ignore the diagonal (topright - bottomleft direction), since those are the variance of the same $X_i$. In the graphical model, $X_1$ is conditionally dependent on $X_4$ given all other variables. Also, $X_2$ is conditionally dependent on $X_3$ given all other variables. We can see that these two relationship rendering high $\Omega_ij$ in the precision matrix, whereas other components are close to 0.

(2) When we look at the "Precision matrix of $X^L$", we can see that the sparsity pattern agrees with the Graphical model. In the graphical model, $X_1$, $X_4$, $X_7$ are in the same clade. We can see that in the precision matrix heatmap, $X_1$, $X_4$, $X_7$ can be clustered due to its relatively high $\Omega_ij$ value. Also, $X_2$, $X_3$, $X_6$ are in the same clade, which is reflected in the precision matrix as well. On the contrary, the leafs that are not in the same clade, especially $X_5$ (sort of an outgroup of the tree) has low $\Omega_ij$ with all other leafs.

## B. Spatial Gaussian Processes.

Consider the data from http://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.0030339, available at https://github.com/stephens999/hgen48600/tree/master/data/CCR5 which you can read into R using code from https://stephens999.github.io/hgen48600/ccr5.html \

Read the data.
```{r}
# load data
ccr5 = read.table("data/CCR5.freq.txt",header=TRUE)
ccr5[,1] = ifelse(ccr5[,1]>180,ccr5[,1]-360,ccr5[,1]) # changes longitudes>180 to negative




```


## B-a) 


